Supplementary Materials: Plastic in the Urban Freshwater Environment

### 2021-12-09 code/analysis_code/exploration.Rmd
Clean up

Quantities of main interest: 
 - particles per liter of microplastics
 - season/sample date
 - population level
 - proximity to water reclamation facility effluent
 - bacteria levels at corresponding samples sites/dates

@@

Nice! Going forward, I should think about whether to keep or remove the watersheds that were only sampled in the last (July) sampling. I also see that some watershed are missing intermediate data points and are not connected. Would we want to connect and skip over a season, or leave them unconnected (looking at Oconee River)

>>>>

Quantities of main interest:

 - particles per liter of microplastics
 - season/sample date
 - population level
 - proximity to water reclamation facility effluent
 - bacteria levels at corresponding samples sites/dates

@@

Going forward, I should think about whether to keep or remove the watersheds that were only sampled in the last (July) sampling. Note: I also see that some watershed are missing intermediate data points and are not connected. Would we want to connect and skip over a season, or leave them unconnected (looking at Oconee River)?

### 2021-12-09 code/analysis_code/subset_modeling.Rmd
Clean up

In this portion of the analysis, we will perform modeling on a smaller subset of data in attempt to improve model performance. The original aim of this document was to perform feature selection/subsetting for model improvement. However, the implementation of the LASSO method eliminated the need for this by essentially automating subset selection. The contents of this document are needed for the final analysis/manuscript product, but are kept in the project repository for understanding of the whole analysis process.

@@

# Model tuning and fitting
Include:
1. Model specification
2. Workflow definition
3. Tuning grid specification
4. Tuning w/ cross-validation + `tune_grid()`

@@

next I want to see what predictors are used in the best lasso model. also need to compare to null model

@@

The results we get from the initial machine learning modeling and from the modeling of this subset data are not great. Let's think of some ways that we can go about improving our model. 
- An option is variable/predictor removal. We tried that with the subset data. We could attempt further subsets going forward. 
- Variable transformation. Some microplastic studies use log transformation on the concentration values to get a normal distribution. 
- Alternatives: account for different distribution. The distribution of microplastic concentration may be considered more Poisson distributed than normally distributed. First, we can test to see whether it falls in either of these distribution categories, and based on the results we can apply a GLM with Poisson distribution. (Tranformation could make intrepretability more difficult - may want to avoid)
- however, regularization may accomplish our goals more efficiently - we are advised that subset selection is not always a great idea and doesn't neccesarily do the best job of addressing overfitting
- could try ridge regression for alternate form of regularization 
- could try decision tree (though we learned these usually don't perform as well as LASSO)

>>>>

In this portion of the analysis, we will perform modeling on a smaller subset of data in attempt to improve model performance. The original aim of this document was to perform feature selection/subsetting for model improvement. However, the implementation of the LASSO method eliminated the need for this by essentially automating subset selection. The contents of this document are *not* needed for the final analysis/manuscript product, but are kept in the project repository for understanding of the whole analysis process.

@@

# Model tuning and fitting
Include:

1. Model specification
2. Workflow definition
3. Tuning grid specification
4. Tuning w/ cross-validation + `tune_grid()`

@@

Next, I want to see what predictors are used in the best lasso model. also need to compare to null model

@@

The results we get from the initial machine learning modeling and from the modeling of this subset data are not great. Let's think of some ways that we can go about improving our model.

- An option is variable/predictor removal. We tried that with the subset data. We could attempt further subsets going forward. 
- Variable transformation. Some microplastic studies use log transformation on the concentration values to get a normal distribution. 
- However, regularization may accomplish our goals more efficiently - we are advised that subset selection is not always a great idea and doesn't neccesarily do the best job of addressing overfitting
- Could try ridge regression for alternate form of regularization 
- Could try decision tree (though we learned these usually don't perform as well as LASSO)

### 2021-12-09 code/processing_code/ARCHIVE/uownprocessing.Rmd
Clean up

@@

#save e.coli.cfu as numeric instead of character
cleanuowndata <- cleanuowndata %>% mutate(e.coli.cfu = as.numeric(e.coli.cfu))
class(cleanuowndata$e.coli.cfu)
```

>>>>

# Merging mass UOWN dataset and July data, which was received later (10/27/21)
```{r}
july <- read.csv(here("data", "raw_data", "uown_summer2021_qtrly.csv"))
glimpse(july)
#The july data also contains 17 columns, just like the mass dataset. This indicates that these datasets should be easy to merge. 
#adjusting class of some variables so that they are compatible between these two datasets before joining
july <- july %>% mutate(ID = as.character(ID), 
               pH = as.numeric(pH)) 
#adjust class of e.coli.cfu from character to numeric to join datasets (and to work with later in analysis)               
uowndata <- uowndata %>% mutate(e.coli.cfu = as.numeric(e.coli.cfu))

uowndata <- full_join(uowndata, july, all = TRUE)
```

@@

### 2021-12-09 products/manuscript/Manuscript.Rmd
Clean up

---
title: "Plastic in the Urban Freshwater Environment"
subtitle: "An Exploratory Study of Microplastics in the Upper Oconee Watershed via Community-Based Efforts"
author: Amelia M. Foley
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
    fig_caption: true
  bookdown::html_document2: 
    toc: false
bibliography: library.bib 
csl: apa.csl
---

@@

Through community-based sample collection and analysis, volunteers quantified microplastics through membrane filtration and visual identification. Preliminary data consistently show that local freshwater microplastic levels are greater than the U.S. average in the Upper Oconee Watershed, a portion of the Oconee River basin with both a high population size and high population density relative to other regions of Georgia. Average and median concentrations of 95 particles/L and 46 particles/L, respectively, were recorded throughout one year of seasonal sampling. Further visual and statistical analysis will confirm the project’s results, and discussion of these results in context will reveal the efficacy of this community-based approach to the quantification of microplastics.

@@

### Present Research
This preliminary study sought to identify the presence of microplastics in the Upper Oconee Watershed, estimate the levels of microplastic pollution present in the freshwater systems of Athens, GA, observe temporal changes and trends, and explore the relationship between microplastic pollution and hypothesized predictors including population level, land use, bacteria level, and wastewater treatment plant proximity (@Murphy2016). Thus, the key questions that we ask in this study are: what are microplastic levels like in and aroudn Athens, GA? Do microplastic levels vary by season? Does local population level, local wastewater treatment, land cover, or general water quality influence microplastic levels? We hypothesized that microplastics would be present in the Upper Oconee Watershed, that levels would be greater in more highly-trafficked areas with greater anthropogenic impact, and that microplastic pollution would correlate to other indicators of water quality [@McCormick2014a].

@@

### Description of study area
The Upper Oconee Watershed is located in Northeast Georgia, spanning from Gainesville on the northern end to Lake Sinclair in Milledgeville, Georgia, on the southern end. The Upper Oconee Watershed encompasses 7590 km2, and includes more than 4000 km of streams [@Fisher]. Land within the Upper Oconee Watershed is historically agricultural, but has been subject to increasing urbanization in the past two decades; the study area contains developed and densely populated urban residential areas [@Cho; @Fisher]. The Oconee Watershed is one of 14 river basins in the state of Georgia; the Oconee River feeds into the Altamaha River, which eventually deposits into the Atlantic Ocean off the Georgia coast (source: UOWN website). The Oconee River begins where the North and Middle Oconee Rivers meet, south of Athens, Georgia. The North and Middle Oconee Rivers begin in Lula, GA, and Brassleton, GA, respectively.

The Upper Oconee Watershed is regularly monitored by the Upper Oconee Watershed Network (UOWN), an Athens-based monitoring, education, advocacy, and recreation organization. Quarterly sampling events occur throughout the year, and the present research project was made possible via collaboration and eventual partnership with UOWN and the quarterly sampling events. The UOWN study area consists of 600 km2.

@@

### UOWN Quarterly Data
This study relies on bublic data from Upper Oconee Watershed Network, Athens, Georgia; retrieved from www.uown.org. UOWN supplied information on samples site names, coordinates, and corresponding watersheds during primary data collection. Following microplastic analysis, UOWN data on the following variables for corresponding samples sites and dates was imported to supplement analysis: E. coli CFU/100mL,  temperature, turbidity, conductivity, and visual score.

@@

## Data Cleaning
The data was cleaned to adjust variable format, calculate microplastic concentration averages, subset sample data and control data, and remove missing observations. Observations with missing location (coordinates, watershed) data were removed, as well as observations which were missing count values for microplastic concentration. UOWN data was cleaned to remove variables with incomplete observations and variables uninformative for analysis. Only variables defined in the study hypothesis were included in the cleaned data. Census data was cleaned and wrangled to match long form FIPS codes with short form FIPS codes. Land cover data was cleaned to extract FIPS codes, to include only data for the state of Georgia, to remove superfluous variables, to format and rename key variables, and to reassign null values. Null land cover proportion values were reassigned to the value zero on the basis that absent values indicated a lack of that particular land cover type in the region (for example, perennial ice/snow was absent in the study region of Georgia).

@@

## Full Analysis
Beyond the bivariate analysis, a full model was built to incorporate all predictors in modeling microplastic concentration. Three modeling techniques were performed in order to perform a multivariate analysis and subset selection: decision tree, LASSO, and random forest. A null model is used to compare model performance and therefore interpret variable importance. The null model results in an RMSE of 109.4. This metric serves as an informative baseline; any additional model should result in a better (lower RMSE) than the null model. The process of building a better model via subset selection and variable regularization reveals important variable relationships and determines variable importance.

@@

## Summary and Interpretation
In this analysis, we examined the distribution of freshwater microplastics in Athens, GA, as well as the relationship between the main outcome, microplastic concentration, and a few key hypothesized predictors: population level, microbial water quality, land cover, and proximity to wastewater treatment. Microplastic levels were monitored over a period of one year, with water samples being collected and processed quarterly in November 2020, February 2021, April 2021, and July 2021. Initial data collection provided sample location data as well as microplastic concentration data. Additional data was later imported, including microbial water quality, local population, land cover, and distance from the nearest water treatment facilities.

Overall, concentration levels were found to be higher than expected based on some previous studies of regional freshwater microplastic pollution. However, the body of research on microplastics as a whole reflects mixed evidence of concentrations. Recorded levels in the literature range from 1 or fewer particles per liter at some sites to greater than 1000 particles per liter at others. There is growing evidence that microplastic pollution in permeates ecosystems in unexpected, supposedly pristine regions across the globe; one study found between 22 and 117 particles per liter in the Antarctic, demonstrating that microplastic particles are being found in places you would never expect.

@@

## Strengths and Limitations
Limitations to the scope of this project include that it was designed as a student-led pilot with minimal funding. Visual identification via dissecting microscope was the method used due to its simplicity and accessibility; visual identification is used in most baseline microplastic studies, but can be up to 70% inaccurate and therefore is usually accompanied by spectroscopy technology which allows chemical identification of suspected microplastic particles and increasing the accuracy of quantification results.

@@

A strength of the study design includes that it incorporates citizen science methods of sample collection and analysis and implements cost-effective procedures. The design of this exploratory analysis makes the study accessible and feasible to reproduce in the field and laboratory. To the knowledge of the authors, this is the first study to examine microplastic pollution in Athens, GA. Therefore, this study accomplishes exploratory aims and meets the growing demand for freshwater microplastics research and monitoring. 
## Conclusions
The present research establishes the existence of microplastics in Athens, GA's freshwater bodies, though there is no clear relationship between microplastic concentration and the hypothesized predictive factors, according to the data recorded and analyzed in this study. Continued research is required in order to identify sources of microplastic pollution so that pollution levels may be mitigated in the future. Areas of focus for the future include furthering temporal, routine data collection, refining quantification methods, and increasing funding for microplastic research technologies.

>>>>

---
title: "Plastic in the Urban Freshwater Environment"
subtitle: "An Exploratory Study of Microplastics in the Upper Oconee Watershed via Community-Based Efforts"
author: Amelia M. Foley
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
    fig_caption: true
  bookdown::html_document2: 
    toc: false
bibliography: library.bib 
csl: elsevier.csl
---

@@

Through community-based sample collection and analysis, volunteers quantified microplastics through membrane filtration and visual identification. Preliminary data consistently show that freshwater microplastics are present in the local Upper Oconee Watershed, a portion of the Oconee River basin with both a high population size and high population density relative to other regions of Georgia. Average and median concentrations of 104 particles/L and 66 particles/L, respectively, were recorded throughout one year of seasonal sampling. Additionally, this project assesses the relationship between microplastic levels and hypothesized predictors of pollution, finding no present correlations between microplastic levels and population, land cover, wastewater treatment, or general water quality. Primarily, this project establishes the presence of microplastics in the region and demonstrates the use of citizen science, community based efforts for advances in science and monitoring, while further studies may continue to investigate predictive factors.

@@

### Present Research
This preliminary study sought to identify the presence of microplastics in the Upper Oconee Watershed, estimate the levels of microplastic pollution present in the freshwater systems of Athens, GA, observe temporal changes and trends, and explore the relationship between microplastic pollution and hypothesized predictors including population level, land use, bacteria level, and wastewater treatment plant proximity (@Murphy2016). Thus, the key questions that we ask in this study are:

 - What are microplastic levels like in and around Athens, GA? 
 - Do microplastic levels vary by season? 
 - Does local population level, local wastewater treatment, land cover, or general water quality influence microplastic levels?

We hypothesized that microplastics would be present in the Upper Oconee Watershed, that levels would be greater in more highly-trafficked areas with greater anthropogenic impact, and that microplastic pollution would correlate to other indicators of water quality [@McCormick2014a].

@@

### Description of study area
The Upper Oconee Watershed is located in Northeast Georgia, spanning from Gainesville on the northern end to Lake Sinclair in Milledgeville, Georgia, on the southern end. The Upper Oconee Watershed encompasses 7590 km^2, and includes more than 4000 km of streams [@Fisher]. Land within the Upper Oconee Watershed is historically agricultural, but has been subject to increasing urbanization in the past two decades; the study area contains developed and densely populated urban residential areas [@Cho; @Fisher]. The Oconee Watershed is one of 14 river basins in the state of Georgia; the Oconee River feeds into the Altamaha River, which eventually deposits into the Atlantic Ocean off the Georgia coast. The Oconee River begins where the North and Middle Oconee Rivers meet, south of Athens, Georgia. The North and Middle Oconee Rivers begin in Lula, GA, and Brassleton, GA, respectively.

The Upper Oconee Watershed is regularly monitored by the Upper Oconee Watershed Network (UOWN), an Athens-based monitoring, education, advocacy, and recreation organization. Quarterly sampling events occur throughout the year, and the present research project was made possible via collaboration and eventual partnership with UOWN and the quarterly sampling events. The UOWN study area consists of 600 km^2.

@@

### UOWN Quarterly Data
This study relies on public data from Upper Oconee Watershed Network, Athens, Georgia; retrieved from www.uown.org. UOWN supplied information on samples site names, coordinates, and corresponding watersheds during primary data collection. Following microplastic analysis, UOWN data on the following variables for corresponding samples sites and dates was imported to supplement analysis: E. coli CFU/100mL,  temperature, turbidity, conductivity, and visual score.

@@

## Data Cleaning
The data was cleaned to adjust variable format, calculate microplastic concentration averages, subset sample data and control data, and remove missing observations. Observations with missing location data (coordinates, watershed) were removed, as well as observations which were missing count values for microplastic concentration. UOWN data was cleaned to remove variables with incomplete observations and variables uninformative for analysis. Only variables defined in the study hypothesis were included in the cleaned data. Census data was cleaned and wrangled to match long form FIPS codes with short form FIPS codes. Land cover data was cleaned to extract FIPS codes, to include only data for the state of Georgia, to remove superfluous variables, to format and rename key variables, and to reassign null values. Null land cover proportion values were reassigned to the value zero on the basis that absent values indicated a lack of that particular land cover type in the region (for example, perennial ice/snow was absent in the study region of Georgia).

@@

## Full Analysis
Beyond the bivariate analysis, a full model was built to incorporate all predictors in modeling microplastic concentration. Three modeling techniques were implemented in order to perform a multivariate analysis and subset selection: decision tree, LASSO, and random forest. A null model is used to compare model performance and therefore interpret variable importance. The null model results in an RMSE of 109.4. This metric serves as an informative baseline; any additional model should result in a better (lower RMSE) than the null model. The process of building a better model via subset selection and variable regularization reveals important variable relationships and determines variable importance.

@@

## Summary and Interpretation
In this analysis, we examined the distribution of freshwater microplastics in Athens, GA, as well as the relationship between the main outcome, microplastic concentration, and a few key hypothesized predictors: population level, microbial water quality, land cover, and proximity to water reclamation facilities. Microplastic levels were monitored over a period of one year, with water samples being collected and processed quarterly in November 2020, February 2021, April 2021, and July 2021. Initial data collection provided sample location data as well as microplastic concentration data. Additional data was later imported, including microbial water quality, local population, land cover, and distance from the nearest water treatment facilities.

Overall, concentration levels were found to be higher than expected based on some previous studies of regional freshwater microplastic pollution. However, the body of research on microplastics as a whole reflects mixed evidence of concentrations. Recorded levels in the literature range from 1 or fewer particles per liter at some sites to greater than 1000 particles per liter at others [@Li2018;@Tibbetts2018]. There is growing evidence that microplastic pollution permeates ecosystems in unexpected, supposedly pristine regions across the globe; one study found between 22 and 117 particles per liter in the Antarctic, demonstrating that microplastic particles are being found in places you would never expect [@Waller2017].

@@

## Strengths and Limitations
Limitations to the scope of this project include that it was designed as a student-led pilot with minimal funding. Visual identification via dissecting microscope was the method used due to its simplicity and accessibility; visual identification is used in most baseline microplastic studies, but can be up to 70% inaccurate and therefore is usually accompanied by spectroscopy technology which allows chemical identification of suspected microplastic particles and increasing the accuracy of quantification results [@Primpke2018].

@@

A strength of the study design includes that it incorporates citizen science methods of sample collection and analysis and implements cost-effective procedures. The design of this exploratory analysis makes the study accessible and feasible to reproduce in the field and laboratory. To the knowledge of the authors, this is the first study to examine microplastic pollution in Athens, GA. Therefore, this study accomplishes exploratory aims and meets the growing demand for freshwater microplastics research and monitoring.

## Conclusions
The present research establishes the existence of microplastics in Athens, GA's freshwater bodies, though there is no clear relationship between microplastic concentration and the hypothesized predictive factors, according to the data recorded and analyzed in this study. Continued research is required in order to identify sources of microplastic pollution so that pollution levels may be mitigated in the future. Areas of focus for the future include furthering  routine and temporal data collection, refining quantification methods, and increasing funding for microplastic research technologies.

### 2021-12-09 products/manuscript/Supplement.Rmd
Clean up

* "**AmeliaFoley-MADA-project.Rproj**" Establishes relative file paths
* "**README.md**" Provides brief order of scripts for reproducing; summarizes project parts
* "**code**" folder contains all code for processing and analysis
  + "**analysis_code**"
    + analysis.Rmd performs linear regression on single predictors and full predictor sets
    + exploration.Rmd contains the exploratory data analysis
    + land_model.Rmd contains statistical models on land cover data
    + modeling.Rmd is the main analysis file containing final models (LASSO)
    + subset_modeling.Rmd is an additional analysis file performing models on subsets of data; it is not used for the manuscript
  + "**processing_code**"
    + censusland_code.Rmd combines land cover data with microplastic data
    + censusprocessing.Rmd imports census tract population data
    + map_images.Rmd combines individual map images to produce a grid
    + mpprocessing.R processes the original microplastic data file and saves cleaned data
    + uownprocessing.Rmd imports additional water quality data from UOWN.org
    + wrfprocessing.Rmd imports distance for nearest water reclamation facility
  + "**readme.md**" provides an order and brief description of dependencies/outputs for processing code files
* "**data**" folder contains all raw and processed data
* "**products**" folder contains files to produce the Manuscript and Supplement
  + "**manuscript**" contains files for producing the manuscript
    + apa.csl optional format for citations
    + elsevier.csl optional format for citations
    + library.bib contains references
    + Manuscript.docx result of knitting the markdown
    + Manuscript.Rmd creates basic manuscript word document
    + style_template.docx placeholder to add special formatting to manuscript
    + Supplement.Rmd creates basic supplemental material word document
    + Supplement.docx result of knitting markdown file
* "**results**" folder contains all plots, tables, figures produced in data analysis

@@

1. mpprocessing.R
2. uownprocessing.Rmd
3. censusprocessing.Rmd
4. censusland_code.Rmd
5. wrfprocessing.Rmd
6. exploration.Rmd

@@

7. analysis.Rmd
8. modeling.Rmd
9. land_model.Rmd
10. map_images.Rmd

@@

11. Manuscript.Rmd
12. Supplement.Rmd

>>>>

* "**AmeliaFoley-MADA-project.Rproj**" Establishes relative file paths
* "**README.md**" Provides brief order of scripts for reproducing; summarizes project parts
* "**code**" folder contains all code for processing and analysis
  + "**analysis_code**"
    + analysis.Rmd performs linear regression on single predictors and full predictor sets
    + exploration.Rmd contains the exploratory data analysis
    + land_model.Rmd contains statistical models on land cover data
    + modeling.Rmd is the main analysis file containing final models (LASSO)
    + subset_modeling.Rmd is an additional analysis file performing models on subsets of data; it is not used for the manuscript
  + "**processing_code**"
    + processing_code.Rmd is the master script for all processing. It combines all of the scripts from below, which have been moved to the `ARCHIVE` folder. 
      + censusland_code.Rmd combines land cover data with microplastic data
      + censusprocessing.Rmd imports census tract population data
      + mpprocessing.R processes the original microplastic data file and saves cleaned data
      + uownprocessing.Rmd imports additional water quality data from UOWN.org
      + wrfprocessing.Rmd imports distance for nearest water reclamation facility
    + map_images.Rmd combines individual map images to produce a grid
  + "**readme.md**" provides an order and brief description of dependencies/outputs for processing code files
* "**data**" folder contains all raw and processed data
* "**products**" folder contains files to produce the Manuscript and Supplement
  + "**manuscript**" contains files for producing the manuscript
    + apa.csl optional format for citations
    + elsevier.csl optional format for citations
    + library.bib contains references
    + Manuscript.docx result of knitting the markdown
    + Manuscript.Rmd creates basic manuscript word document
    + style_template.docx placeholder to add special formatting to manuscript
    + Supplement.Rmd creates basic supplemental material word document
    + Supplement.docx result of knitting markdown file
* "**results**" folder contains all plots, tables, figures produced in data analysis

@@

1. processing_code.Rmd

@@

2. analysis.Rmd
3. modeling.Rmd
4. land_model.Rmd

5. map_images.Rmd

@@

6. Manuscript.Rmd
7. Supplement.Rmd
